\bf A Unitary RG-based Auxiliary Model Approach to Strongly-Correlated Electrons

### 2025-05-02 TilingProjectDraft/tilingProject.tex
latest version of manuscript and notes, with most of the figures added

\subsection{Unitary RG analysis of the embedded eSIAM}
We studied the low-energy physics of this model through a unitary renormalisation group calculation that decouples the high-energy modes of the conduction bath and incorporates their effects in the form of renormalised Hamiltonian couplings. We find that the bath interaction \(W\) remains marginal (does not undergo any renormalisation), while the renormalisation in the momentum-resolved Kondo coupling \(J^{(j)}\) at the \(j^\text{th}\) step is of the form
\begin{equation}\begin{aligned}\label{KondoRGequation}
	\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2} = -\sum_{{\bf q} \in \text{PS}} \frac{J^{(j)}_{{\bf k}_2,{\bf q}} J^{(j)}_{{\bf q},{\bf k}_1} + 4J^{(j)}_{{\bf q}, {\bf \bar q}} W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}}{\omega - \frac{1}{2}|\varepsilon_j| + J^{(j)}_{{\bf q}}/4 + W_{{\bf q}}/2}~,
\end{aligned}\end{equation}
where \(\varepsilon_j\) is the energy of the shell being decoupled at the \(j^\text{th}\) step, and the sum is over all momentum states \({\bf q}\) in the particle sector (PS) of the shell \(\varepsilon_j\) (that is, all states that are occupied at \(T=0\) and in the absence of any quantum fluctuations). The Kondo coupling and the RG equation as well have certain symmetries under transformations in the Brillouin zone. If any one of the momenta \({\bf k}_1\) or \({\bf k}_2\) are translated by \(\pi\) (\(k_{1x} \to k_{1x} + \pi, k_{1y} \to k_{1y}+\pi\)), the coupling as well as the RG equation changes sign. Translating both the momenta leads to the reversal of the sign change. 
\begin{equation}\begin{aligned}\label{hamiltonanSymms}
	\Delta J^{(j)}_{{\bf k}_1 + \boldsymbol{\pi}, {\bf k}_2} = \Delta J^{(j)}_{{\bf k}_1, {\bf k}_2 + \boldsymbol{\pi}} = -\Delta J^{(j)}_{{\bf k}_1 + \boldsymbol{\pi}, {\bf k}_2 + \boldsymbol{\pi}} = -\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2}~.
\end{aligned}\end{equation}
These transformations involve translating one or both momenta through the center of the Brillouin zone.

@@

Coleman and Ioffe have shown that by mapping the two-channel Kondo problem to a ``compactified" form where the impurity spin hybridises with the spin and isospin degrees of freedom of a single conduction channel, the impurity self-energy can be shown to display marginal Fermi liquid behaviour~\cite{Coleman_tsvelik,Schofield1997}:
\begin{equation}\begin{aligned}
	\Sigma \propto \omega \ln \omega~.
\end{aligned}\end{equation}
Some of us have shown in a previous work~\cite{Patra2023MCK} that this can also be obtained from a renormalisation group calculation of the low-energy excitations of the two-channel overscreened fixed point. For the sake of completeness, we have reproduced the calculation in Appendix~\ref{mflDerivation}.

@@

In order to extract the singular, critical part of the Hamiltonian, we focus on the \({\bf q}_1 = {\bf q}_2\) forward scattering processes:
\begin{equation}\begin{aligned}\label{HKModel}
	\Delta \tilde H_{{\bf q}_1 = {\bf q}_2} = \sum_{{\bf q},\sigma}\epsilon_{{\bf q}}{r}_{{\bf q},\sigma} + u\sum_{{\bf q}, \sigma}r_{{\bf q} \sigma} r_{{\bf q} \bar\sigma}
\end{aligned}\end{equation}
where \(\epsilon_{\bf q}\) and \(u\) are emergent energy scales:
\begin{equation}\begin{aligned}
	\epsilon_{\bf q} = \text{sign}\left(\varepsilon_{{\bf N}_1 + {\bf q}}\right)\frac{\varepsilon_{{\bf N}_1 + {\bf q}}^2}{-W} + \frac{{J^*}^2}{-4W},~u = \frac{{J^*}^2}{4W}~.
\end{aligned}\end{equation}
Eq.~\ref{HKModel} describes a Hatsugai-Kohmoto model~\cite{Baskaran1991,Hatsugai1992}, an exactly solvable model of correlated electrons; by tuning the ratio of the interaction strength \(u\) and the bandwidth \(w\), the model describes a transition between a non-Fermi liquid and a Mott insulator. The non-Fermi liquid nature of the gapless excitations in the metallic phase of the model manifests in the form of a divergent one-particle self-energy at the non-interacting Fermi surface~\cite{Phillips2020}:
\begin{equation}\begin{aligned}
	\Sigma_{{\bf q}}(\omega) = -\frac{u^2/4}{\omega - \epsilon_{\bf q}}, ~\Sigma_{\epsilon_{\bf q}=0}(\omega \to 0)  \to \infty~.
\end{aligned}\end{equation}
Beyond the transition and inside the Mott insulating phase, this zero frequency pole continues to exist and marks the hard gap in the spectral function for charge excitations.

@@

In order to better understand the nature of the metal in the partially gapped region, we have also calculated the \(k-\)space self-energy \(\tilde \Sigma({\bf k}(\omega)\), using Dyson's equation:
\begin{equation}\begin{aligned}
	\tilde \Sigma({\bf k}(\omega) = 1/\tilde G^{(0)}_{\bf k}(\omega) - 1/\tilde G_{\bf k}(\omega)~,
\end{aligned}\end{equation}
where \(\tilde G_k\) is the interacting \(k-\)space Greens function (whose spectral function is shown in Fig.~\ref{tiledSpecFunc}) while \(\tilde G_k^{(0)}\) is the non-interacting \(W=0\) Greens function. We show the zero frequency value of the self-energy in Fig.~\ref{selfenergy}, and the partial gapping of the Fermi surface is seen to arise from the appearance of poles in the self-energy at the antinodal region. As the partial gap expands, more and more \(k-\)space points exhibit poles in their self-energy. This, along with the behaviour of the spectral function, indicates that in the pseudogap region, the disconnected Fermi surface coexists with a Luttinger surface.
\begin{figure}[htpb]
	\centering
	\includegraphics[width=0.23\textwidth]{kspaceDOS-1.pdf}
	\includegraphics[width=0.23\textwidth]{kspaceDOS-2.pdf}
	\includegraphics[width=0.23\textwidth]{kspaceDOS-3.pdf}
	\includegraphics[width=0.23\textwidth]{kspaceDOS-4.pdf}
	\caption{One-particle \(k-\)space spectral function. In the absence of bath interactions, all points on Fermi surface are gapless (first panel). Proceeding through the pseudogap results in the gapping out of Fermi surface points starting from the antinode (second panel) and approaching the node (third panel), until finally only the node remains at the critical point (fourth panel).}
	\label{tiledSpecFunc}
\end{figure}
\begin{figure}[htpb]
	\centering
	\includegraphics[width=0.23\textwidth]{selfEnergyKspace-1.pdf}
	\includegraphics[width=0.23\textwidth]{selfEnergyKspace-2.pdf}
	\includegraphics[width=0.23\textwidth]{selfEnergyKspace-3.pdf}
	\includegraphics[width=0.23\textwidth]{selfEnergyKspace-4.pdf}
	\caption{\(k-\)space self-energy. For \(W=0\), the self-energy is vanishingly small, leading to a Fermi liquid phase described by well-defined quasiparticles. Inside the pseudogap, the self-energy diverges near the antinodes, leading to their gapping. The coexistence of gapped and gapless points on the Fermi surface in the pseudogapped region leads to non-Fermi liquid behaviour of the gapless excitations, as captured by the emergent two-channel behaviour within the impurity model. The metal at the transition is supported only the four nodal points.}
	\label{selfenergy}
\end{figure}

@@

@@

The renormalisation in \(R\) leads to a renormalisation in the single-particle self-energy~\cite{anirbanmott1}. The \(k-\)space-averaged self-energy renormalisation is
\begin{eqnarray}
	\Delta \Sigma(\omega) = \rho {R^*}^2\int_0^{\epsilon^*} \frac{d\epsilon_j}{\omega - \epsilon_j/2 + R_j/8}~.
\end{eqnarray}
The density of states can be approximated to be \(N^*/R^*\), where \(N^*\) is the total number of states over the interval \(R^*\). As suggested by the fixed point value of \(R_j\), we can approximate its behaviour near the fixed point by a linear dependence on the dispersion \(\epsilon_j\). The two limits of the integration are the starting and ending points of the RG. We start the RG very close to the Fermi surface and move towards the fixed point \(\epsilon^*\). Near the starting point, we substitute \(\epsilon = 0\) and \(R = \omega\), following the fixed point condition. From the fixed point condition, we also substitute \(R^*/8 = \omega - \frac{1}{2}\epsilon^*\). On defining \(\bar \omega = N^* \left(\omega - \frac{1}{2}\epsilon^*\right)\), we can write
\begin{eqnarray}
	\label{self_energy}
	\Delta \Sigma(\omega) \sim  \bar \omega \ln \frac{N^* \omega}{\bar \omega}~.
\end{eqnarray}
The self-energy also provides the quasiparticle residue for each channel\cite{anirbanmott1}:
\begin{eqnarray}
	Z(\bar\omega) = \left(2 - \ln \frac{2\bar\omega}{N^* \omega}\right) ^{-1}~.
\end{eqnarray}
As \(\omega \to 0\), the \(Z\) vanishes, implying that the ground state is {\it not adiabatically connected} to the Fermi gas in the presence of the NFL terms.
This is the orthogonality catastrophe~\cite{varma2002singular,anderson_infraredcat,yamada_catastrophe,yamada1979orthogonality} in the two-channel Kondo problem, and it is brought about by the presence of the channel-non diagonal terms in Eq.~\ref{mfl_large}.

>>>>

\subsection{Unitary RG analysis of the embedded eSIAM}
We studied the low-energy physics of this model through a unitary renormalisation group calculation that decouples the high-energy modes of the conduction bath and incorporates their effects in the form of renormalised Hamiltonian couplings. We find that the bath interaction \(W\) remains marginal (does not undergo any renormalisation), while the renormalisation in the momentum-resolved Kondo coupling \(J^{(j)}\) at the \(j^\text{th}\) step is of the form
\begin{equation}\begin{aligned}\label{KondoRGequation}
	\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2} = -\sum_{{\bf q} \in \text{PS}} \frac{J^{(j)}_{{\bf k}_2,{\bf q}} J^{(j)}_{{\bf q},{\bf k}_1} + 4J^{(j)}_{{\bf q}, {\bf \bar q}} W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}}{\omega - \frac{1}{2}|\varepsilon_j| + J^{(j)}_{{\bf q}}/4 + W_{{\bf q}}/2}~,
\end{aligned}\end{equation}
where \(\varepsilon_j\) is the energy of the shell being decoupled at the \(j^\text{th}\) step, and the sum is over all momentum states \({\bf q}\) in the particle sector (PS) of the shell \(\varepsilon_j\) (that is, all states that are occupied at \(T=0\) and in the absence of any quantum fluctuations). The Kondo coupling and the RG equation as well have certain symmetries under transformations in the Brillouin zone. If any one of the momenta \({\bf k}_1\) or \({\bf k}_2\) are translated by \(\pi\) (\(k_{1x} \to k_{1x} + \pi, k_{1y} \to k_{1y}+\pi\)), the coupling as well as the RG equation changes sign. Translating both the momenta leads to the reversal of the sign change. 
\begin{equation}\begin{aligned}\label{hamiltonanSymms}
	\Delta J^{(j)}_{{\bf k}_1 + \boldsymbol{\pi}, {\bf k}_2} = \Delta J^{(j)}_{{\bf k}_1, {\bf k}_2 + \boldsymbol{\pi}} = -\Delta J^{(j)}_{{\bf k}_1 + \boldsymbol{\pi}, {\bf k}_2 + \boldsymbol{\pi}} = -\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2}~.
\end{aligned}\end{equation}
These transformations involve translating one or both momenta through the center of the Brillouin zone. \textcolor{red}{Is this a d-wave symmetry of the RG equation?}

@@

Coleman and Ioffe have shown that by mapping the two-channel Kondo problem to a ``compactified" form where the impurity spin hybridises with the spin and isospin degrees of freedom of a single conduction channel, the impurity self-energy \(\Sigma = \Sigma^\prime + i\Sigma^{\prime\prime}\) can be shown to display marginal Fermi liquid behaviour~\cite{Coleman_tsvelik,Schofield1997}:
\begin{equation}\begin{aligned}
	\Sigma^\prime \propto \omega \ln \omega, ~\Sigma^{\prime\prime} \propto \omega~.
\end{aligned}\end{equation}
Some of us have shown in a previous work~\cite{Patra2023MCK} that this can also be obtained from a renormalisation group calculation of the low-energy excitations of the two-channel overscreened fixed point. For the sake of completeness, we have reproduced the calculation in Appendix~\ref{mflDerivation}.

@@

In order to extract the singular, critical part of the Hamiltonian, we focus on the \({\bf q}_1 = {\bf q}_2\) forward scattering processes:
\begin{equation}\begin{aligned}\label{HKModel}
	\Delta \tilde H_{{\bf q}_1 = {\bf q}_2} = \sum_{{\bf q},\sigma}\epsilon_{{\bf q}}{r}_{{\bf q},\sigma} + u\sum_{{\bf q}, \sigma}r_{{\bf q} \sigma} r_{{\bf q} \bar\sigma}
\end{aligned}\end{equation}
where \(\epsilon_{\bf q}\) and \(u\) are emergent energy scales:
\begin{equation}\begin{aligned}
	\epsilon_{\bf q} = \text{sign}\left(\varepsilon_{{\bf N}_1 + {\bf q}}\right)\frac{\varepsilon_{{\bf N}_1 + {\bf q}}^2}{-W} + \frac{{J^*}^2}{-4W},~u = \frac{{J^*}^2}{4W}~.
\end{aligned}\end{equation}
Eq.~\ref{HKModel} describes a Hatsugai-Kohmoto model~\cite{Baskaran1991,Hatsugai1992}, an exactly solvable model of correlated electrons; by tuning the ratio of the interaction strength \(u\) and the bandwidth \(w\), the model describes a transition between a non-Fermi liquid and a Mott insulator. The non-Fermi liquid nature of the gapless excitations in the metallic phase of the model manifests in the form of a divergent one-particle self-energy at the non-interacting Fermi surface~\cite{Phillips2020}:
\begin{equation}\begin{aligned}
	\Sigma_{{\bf q}}(\omega) = -\frac{u^2/4}{\omega - \epsilon_{\bf q}}, ~\Sigma_{\epsilon_{\bf q}=0}(\omega \to 0)  \to \infty~,
\end{aligned}\end{equation}
while the spectral function $A(k,\omega=0)$ shows a sharp peak at each of the four nodal Fermi surface points. This will be observed to lead to a nodal non-Fermi liquid metal shown below in Fig.\ref{tiledSpecFunc} (fourth panel). Beyond the transition and inside the Mott insulating phase, this zero frequency pole continues to exist and marks the hard gap in the spectral function for charge excitations. We note that the ${\bf q}_{1}\neq {\bf q}_{2}$ scattering processes correspond to the forward and tangential scattering of the coherent superposition of \({\bf N}_1 = \left(\pi/2, \pi/2\right)\) and its nested partner \({\bf N}_1 + {\bf Q}_1 = \left(-\pi/2, -\pi/2\right)\) along the respective Fermi arcs. Such scattering processes add non-singular contributions to the self-energy, and do not open gaps at the Fermi surface. WE EXPECT SUCH A NODAL NON-FERMI LIQUID METAL TO EXHIBIT VOLUME LAW SCALING OF THE BIPARTITION ENTANGLEMENT ENTROPY.

@@

In order to better understand the nature of the metal in the partially gapped region, we have also calculated the \(k-\)space self-energy \(\tilde \Sigma({\bf k}(\omega)\), using Dyson's equation:
\begin{equation}\begin{aligned}
	\tilde \Sigma({\bf k}(\omega) = 1/\tilde G^{(0)}_{\bf k}(\omega) - 1/\tilde G_{\bf k}(\omega)~,
\end{aligned}\end{equation}
where \(\tilde G_k\) is the interacting \(k-\)space Greens function (whose spectral function is shown in Fig.~\ref{tiledSpecFunc}) while \(\tilde G_k^{(0)}\) is the non-interacting \(W=0\) Greens function. We show the zero frequency value of the self-energy in Fig.~\ref{selfenergy}, and the partial gapping of the Fermi surface is seen to arise from the appearance of poles in the self-energy at the antinodal region. As the partial gap expands, more and more \(k-\)space points exhibit poles in their self-energy. This, along with the behaviour of the spectral function, indicates that in the pseudogap region, the disconnected Fermi surface coexists with a Luttinger surface. WE EXPECT THAT THE JOURNEY THROUGH THE PSEUDOGAP DOES NOT VIOLATE LUTTINGER'S THEOREM; THIS CAN BE SEEN BY COUNTING BOTH THE POLES AND THE ZEROS OF THE SINGLE PARTICLE GREENS FUNCTION. CHECK.
\begin{figure}[htpb]
	\centering
	\includegraphics[width=0.23\textwidth]{kspaceDOS-1.pdf}
	\includegraphics[width=0.23\textwidth]{kspaceDOS-2.pdf}
	\includegraphics[width=0.23\textwidth]{kspaceDOS-3.pdf}
	\includegraphics[width=0.23\textwidth]{kspaceDOS-4.pdf}
	\caption{One-particle \(k-\)space spectral function. In the absence of bath interactions, all points on Fermi surface are gapless (first panel). Proceeding through the pseudogap results in the gapping out of Fermi surface points starting from the antinode (second panel) and approaching the node (third panel), until finally only the node remains at the critical point (fourth panel).}
	\label{tiledSpecFunc}
\end{figure}
\begin{figure}[htpb]
	\centering
	\includegraphics[width=0.23\textwidth]{selfEnergyKspace-1.pdf}
	\includegraphics[width=0.23\textwidth]{selfEnergyKspace-2.pdf}
	\includegraphics[width=0.23\textwidth]{selfEnergyKspace-3.pdf}
	\includegraphics[width=0.23\textwidth]{selfEnergyKspace-4.pdf}
	\caption{\(k-\)space self-energy. For \(W=0\), the self-energy is vanishingly small, leading to a Fermi liquid phase described by well-defined quasiparticles. Inside the pseudogap, the self-energy diverges near the antinodes, leading to their gapping. The coexistence of gapped and gapless points on the Fermi surface in the pseudogapped region leads to non-Fermi liquid behaviour of the gapless excitations, as captured by the emergent two-channel behaviour within the impurity model. The metal at the transition is supported only the four nodal points.}
	\label{selfenergy}
\end{figure}

@@

In order to probe how the metal becomes increasingly correlated as it is tuned through the pseudogap, we compare the quasiparticle residue \(Z_\text{imp}\) and the scattering rate \(\Gamma\) of the quasiparticles, at the beginning and end of the pseudogap. For the regime \(|W| \gtrapprox |W|^*\), the low-energy excitations obey the marginal Fermi liquid theory: \(\Sigma^{\prime} \sim \omega\ln\omega, \Sigma^{\prime\prime}\sim-\omega\). This gives rise to a linear scattering rate \(\Gamma \sim \omega\) (arising from the imaginary part of the self-energy), and a logarithmically vanishing quasiparticle residue~\cite{varma2002singular}
\begin{equation}\begin{aligned}
Z_\text{imp}(\omega) = (1 - \frac{d\Sigma^\prime}{d\omega})^{-1} \sim (c - \ln \omega)^{-1}
\end{aligned}\end{equation}
where \(c\) is a constant. Close to the transition (\(|W| \lesssim |W_c|\)), the low-energy theory is described by a Hatsugai-Kohmoto model, with a diverging self-energy: 
\begin{equation}\begin{aligned}
	\Sigma({\bf k}) = U^2/\left(\omega - \epsilon_{\bf k}\right) = U^2 \left[\mathcal{P}(\frac{1}{\omega - \epsilon_{\bf k}}) - i\pi \delta(\omega - \epsilon_{\bf k})\right]~,
\end{aligned}\end{equation}
where \(\mathcal{P}\left(\cdot\right)\) represents the principal value, and we used the Sokhotski–Plemelj theorem to write down the final expression. For small but non-zero values of \(\omega - \epsilon_{\bf k}\), we get \(Z_\text{imp} \sim \omega^2/U^2 \). The metal near the transition therefore displays a much more rapid vanishing of the quasiparticle residue compared to the logarithmic vanishing for the marginal Fermi liquid. The scattering rate precisely at the surface \(\omega=\epsilon_{\bf k}\), in fact, diverges: \(\Gamma \sim U^2\delta(\omega - \epsilon_{\bf k})\), in contrast to the linearly vanishing scattering rate of the MFL.

@@

The renormalisation in \(R\) leads to a renormalisation in the real part of the single-particle self-energy~\cite{anirbanmott1}. The \(k-\)space-averaged self-energy renormalisation is
\begin{eqnarray}
	\Delta \Sigma^\prime(\omega) = \rho {R^*}^2\int_0^{\epsilon^*} \frac{d\epsilon_j}{\omega - \epsilon_j/2 + R_j/8}~.
\end{eqnarray}
The density of states can be approximated to be \(N^*/R^*\), where \(N^*\) is the total number of states over the interval \(R^*\). As suggested by the fixed point value of \(R_j\), we can approximate its behaviour near the fixed point by a linear dependence on the dispersion \(\epsilon_j\). The two limits of the integration are the starting and ending points of the RG. We start the RG very close to the Fermi surface and move towards the fixed point \(\epsilon^*\). Near the starting point, we substitute \(\epsilon = 0\) and \(R = \omega\), following the fixed point condition. From the fixed point condition, we also substitute \(R^*/8 = \omega - \frac{1}{2}\epsilon^*\). On defining \(\bar \omega = N^* \left(\omega - \frac{1}{2}\epsilon^*\right)\), we can write
\begin{eqnarray}
	\label{self_energy}
	\Delta \Sigma^\prime(\omega) \sim  \bar \omega \ln \frac{N^* \omega}{\bar \omega}~.
\end{eqnarray}
The imaginary part of the self-energy can be obtained using the Kramers-Kronig relations:
\begin{equation}\begin{aligned}
\Delta \Sigma^{\prime\prime}(\omega) = \frac{1}{\pi}\mathcal{P} \int_{-\infty}^\infty d\omega^\prime \frac{\Sigma^\prime(\omega^\prime)}{\omega - \omega^\prime} \sim -\bar\omega~.
\end{aligned}\end{equation}

The self-energy also provides the quasiparticle residue for each channel\cite{anirbanmott1}:
\begin{eqnarray}
	Z(\bar\omega) = \left(2 - \ln \frac{2\bar\omega}{N^* \omega}\right) ^{-1}~.
\end{eqnarray}
As \(\omega \to 0\), the \(Z\) vanishes, implying that the ground state is {\it not adiabatically connected} to the Fermi gas in the presence of the NFL terms.
This is the orthogonality catastrophe~\cite{varma2002singular,anderson_infraredcat,yamada_catastrophe,yamada1979orthogonality} in the two-channel Kondo problem, and it is brought about by the presence of the channel-non diagonal terms in Eq.~\ref{mfl_large}.
